---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tuikr

<!-- badges: start -->
[![DOI](https://zenodo.org/badge/313863336.svg)](https://zenodo.org/badge/latestdoi/313863336)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![R-CMD-check](https://github.com/emraher/tuikr/workflows/R-CMD-check/badge.svg)](https://github.com/emraher/tuikr)
[![pkgdown](https://github.com/emraher/tuikr/actions/workflows/pkgdown.yaml/badge.svg)](https://emraher.github.io/tuikr/)
<!-- badges: end -->

R package for accessing Turkish Statistical Institute (TUIK) data from two portals:

- **Statistical data**: Themes, file downloads, SDMX dataflows, and legacy databases from [veriportali.tuik.gov.tr](https://veriportali.tuik.gov.tr/)
- **Geographic data**: Maps and spatial statistics from [cip.tuik.gov.tr](https://cip.tuik.gov.tr/)

## Installation

```r
# Install from GitHub
devtools::install_github("emraher/tuikr")
```

## Quick Start

### Statistical Data

```{r statistical-example, eval = identical(Sys.getenv("IN_PKGDOWN"), "true")}
library(tuikr)

# 1. List themes
themes <- statistical_themes()

# 2. Tables for Population & Demography (theme 11)
pop_tables <- statistical_tables("11")

# 3. SDMX dataflows only
pop_flows <- dplyr::filter(pop_tables, node_type == "dataflow")

# 4. File downloads expose a direct table_url
pop_files <- dplyr::filter(pop_tables, node_type == "istab")
pop_files$table_url[1]

# 5. Download one dataset
pop_data <- statistical_data(pop_flows$dataflow_id[1])

# 6. Legacy database URLs
pop_dbs <- statistical_databases("11")

# 7. All portal resources
pop_resources <- statistical_resources("11")
```

`statistical_data()` adds adjacent `*_label` columns when TUIK exposes
human-readable code-list metadata. The default `key = "ALL"` works for many
datasets, but some SDMX dataflows need a narrower key to constrain the
remaining dimensions.

### Geographic Data

```{r geographic-example, eval = identical(Sys.getenv("IN_PKGDOWN"), "true")}
library(tuikr)

# List available geographic variables
variables <- geo_data()
head(variables, 3)

# Download data for a specific variable
population <- geo_data(
  var_num = "ADNKS-GK137473-O29001",
  var_level = 3,
  var_source = "medas",
  var_period = "yillik",
  var_recordnum = 5
)
head(population, 3)

# Get map boundaries at different levels
nuts2_map <- geo_map(level = 2)  # 26 regions
nuts3_map <- geo_map(level = 3)  # 81 provinces
lau1_map <- geo_map(level = 4)   # 973 districts
settlements <- geo_map(level = 9)  # settlement points

# Preview map data
head(nuts3_map, 3)
```

`geo_map()` returns `sf` objects in WGS 84 (EPSG:4326), ready for
`dplyr::left_join()` on the `code` column when you want to combine boundaries
with values returned by `geo_data()`.

## Available Map Levels

| Level | Geography | Count | Geometry |
|------:|:----------|------:|:---------|
| 2 | NUTS-2 regions | 26 | MULTIPOLYGON |
| 3 | NUTS-3 provinces | 81 | MULTIPOLYGON |
| 4 | LAU-1 districts | 973 | MULTIPOLYGON |
| 9 | Settlement points | 1,003 | POINT |

## Vignettes

- [Getting Started](https://github.com/emraher/tuikr/blob/dev/vignettes/getting-started.Rmd)
- [Geographic Mapping](https://github.com/emraher/tuikr/blob/dev/vignettes/geographic-mapping.Rmd)

## License

MIT © Emrah Er
