---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  cache = TRUE
)
```

# tuikr

<!-- badges: start -->
[![DOI](https://zenodo.org/badge/313863336.svg)](https://zenodo.org/badge/latestdoi/313863336)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/emraher/tuikr/workflows/R-CMD-check/badge.svg)](https://github.com/emraher/tuikr)
<!-- badges: end -->

**tuikr** provides functions to query Turkish Statistical Institute (TUIK) data from two portals:

- **Statistical data**: Themes, file downloads, SDMX dataflows, and legacy databases from [veriportali.tuik.gov.tr](https://veriportali.tuik.gov.tr/)
- **Geographic data**: Maps and spatial statistics from [cip.tuik.gov.tr](https://cip.tuik.gov.tr/)

**Disclaimer**: This package is not affiliated with, endorsed by, or connected to Turkish Statistical Institute (TUIK). It is an independent tool for academic research purposes.

## Installation

```r
# Install from GitHub
devtools::install_github("emraher/tuikr")
```

## Quick Start

### Statistical Data

#### 1. List themes
```{r statistical-example1, eval = TRUE}
library(tuikr)
library(tibble)

(theme_catalog <- statistical_themes())
```

#### 2. Tables for Population & Demography (theme 11)
```{r statistical-example2, eval = TRUE}
(population_tables <- statistical_tables("11"))
```

#### 3. SDMX dataflows only
```{r statistical-example3, eval = TRUE}
(population_dataflows <- dplyr::filter(
  population_tables,
  node_type == "dataflow"
))
```

#### 4. File downloads expose a direct table_url
```{r statistical-example4, eval = TRUE}
population_files <- dplyr::filter(
  population_tables,
  node_type == "istab"
)
population_files$table_url[1]
```

#### 5. Download one dataset
```{r statistical-example5, eval = TRUE}
(population_observations <- statistical_data(
  population_dataflows$dataflow_id[1]
))
```

`statistical_data()` adds adjacent `*_label` columns when TUIK exposes
human-readable code-list metadata. The default `key = "ALL"` works for many
datasets, but some SDMX dataflows need a narrower key to constrain the
remaining dimensions.

#### 6. Legacy database URLs
```{r statistical-example6, eval = TRUE}
(population_databases <- statistical_databases("11"))
```

#### 7. All portal resources
```{r statistical-example7, eval = TRUE}
(population_resources <- statistical_resources("11"))
```

#### 8. Press releases and reports keep their portal URLs
```{r statistical-example8, eval = TRUE}
population_publications <- dplyr::filter(
  population_resources,
  resource_type %in% c("press", "report")
)
population_publications |>
  dplyr::select(resource_type, resource_name, resource_url) |>
  dplyr::slice(3)
```



### Geographic Data

#### 1. List available geographic variables

```{r geographic-example1, eval = TRUE}
(geo_variable_catalog <- geo_data())
```

#### 2. List geographic variables in Turkish

```{r geographic-example2, eval = TRUE}
(geo_variable_catalog_tr <- geo_data(lang = "tr"))
```

#### 3. Download data for a specific variable

```{r geographic-example3, eval = TRUE}
(population_values <- geo_data(
  var_num = "ADNKS-GK137473-O29001",
  var_level = 3
))
```

#### 4. Get map boundaries at different levels

```{r geographic-example4, eval = TRUE}
nuts2_map <- geo_map(level = 2)  # 26 regions
nuts3_map <- geo_map(level = 3)  # 81 provinces
lau1_map <- geo_map(level = 4)   # 973 districts
settlements <- geo_map(level = 9)  # settlement points
```

`geo_map()` returns `sf` objects in WGS 84 (EPSG:4326), ready for
`dplyr::left_join()` on the `code` column when you want to combine boundaries
with values returned by `geo_data()`.



## MCP Server

The repository also ships a [FastMCP](https://gofastmcp.com) server
(`mcp-server/`) that exposes the same TUIK data surfaces — portal discovery,
SDMX downloads, and geographic data — as Model Context Protocol tools for
Claude Desktop, Claude Code, and other MCP clients. See
[mcp-server/README.md](mcp-server/README.md) for installation and usage.

## Learn More

- **[Getting Started](https://emraher.github.io/tuikr/articles/getting-started.html)**
- **[Geographic Mapping](https://emraher.github.io/tuikr/articles/geographic-mapping.html)**
- **[Function Reference](https://emraher.github.io/tuikr/reference/index.html)**
