---
title: "Downloading and Plotting Maps with tuikr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Downloading and Plotting Maps with tuikr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
build_examples <- identical(Sys.getenv("IN_PKGDOWN"), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = build_examples,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)
```

## Introduction

`tuikr` provides both the indicator values and the ready-to-join geometries you
need for a map. The workflow is:

1. discover a geographic variable with `tuikr::geo_data()`
2. confirm the available NUTS levels in `var_levels`
3. download the values for one level
4. join them to `tuikr::geo_map()`
5. plot the result

## Find a Geographic Series

Start in metadata mode and inspect the housing-sales indicator used in the map
below.

```{r geo-metadata}
geo_variable_catalog <- tuikr::geo_data()
geo_variable_catalog_tr <- tuikr::geo_data(lang = "tr")

housing_sales_series <- geo_variable_catalog |>
  dplyr::filter(var_num == "INS-GK055-O006")

housing_sales_series |>
  dplyr::select(
    var_name,
    var_num,
    var_levels,
    var_period,
    var_source,
    var_recordnum
  )

geo_variable_catalog_tr |>
  dplyr::filter(var_num == "INS-GK055-O006") |>
  dplyr::select(var_name, var_num)
```

Use `lang = "tr"` if you want Turkish indicator names in the metadata table or
in the downloaded value column.

`var_levels` is a list-column. Pull the first element to see the supported
levels for this series.

```{r geo-levels}
housing_sales_series$var_levels[[1]]
```

## Download the Values

`INS-GK055-O006` is published at level 3, which is the provincial map.

```{r geo-download}
housing_sales <- tuikr::geo_data(
  var_num = "INS-GK055-O006",
  var_level = 3,
  var_source = "ilGostergeleri",
  var_period = "yillik",
  var_recordnum = 5
) |>
  dplyr::mutate(
    date = as.integer(date),
    konut_satis_sayilari_toplam = as.numeric(konut_satis_sayilari_toplam)
  )

housing_sales |>
  dplyr::slice_head(n = 8)
```

## Download the Boundaries

`tuikr::geo_map(3)` returns an `sf` object that is already keyed by the `code`
column used in `tuikr::geo_data()`.

```{r geo-boundaries}
province_boundaries <- tuikr::geo_map(3)

province_boundaries |>
  sf::st_drop_geometry() |>
  dplyr::select(code, ad) |>
  dplyr::slice_head(n = 6)
```

If you only want the attribute table, call `tuikr::geo_map(level = 3, dataframe = TRUE)`.

## Join and Plot

Filter to the latest available year, join on `code`, and plot a choropleth with
`ggplot2::geom_sf()`.

```{r geo-plot}
latest_year <- max(housing_sales$date, na.rm = TRUE)

housing_sales_latest <- housing_sales |>
  dplyr::filter(date == latest_year)

housing_sales_map <- province_boundaries |>
  dplyr::left_join(housing_sales_latest, by = "code")

ggplot2::ggplot(housing_sales_map) +
  ggplot2::geom_sf(
    ggplot2::aes(fill = konut_satis_sayilari_toplam),
    color = "white",
    linewidth = 0.1
  ) +
  ggplot2::coord_sf(datum = NA) +
  ggplot2::scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "bottom",
    legend.key.width = grid::unit(3, "cm")
  ) +
  ggplot2::labs(
    title = sprintf("Housing Sales by Province (%s)", latest_year),
    fill = "Sales",
    caption = "Source: TUIK"
  )
```

## Troubleshooting

If a request fails, go back to the metadata table and inspect `var_levels`
before changing `var_level`. Many indicators are not available at every
geographic level.

If the join produces missing fills, inspect the boundary keys with
`sf::st_drop_geometry(province_boundaries)` or call `tuikr::geo_map(level = 3, dataframe = TRUE)`
to compare codes directly.

If you need a different geometry type, remember that levels 2, 3, and 4 return
polygon boundaries while level 9 returns settlement points.
