---
title: "Geographic Mapping with tuikr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Geographic Mapping with tuikr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
build_examples <- identical(Sys.getenv("IN_PKGDOWN"), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = build_examples,
  fig.width = 8,
  fig.height = 6
)
```

## Introduction

Use `tuikr` to discover TUIK geographic variables, download values by NUTS
level, and join them to ready-made `sf` geometries for mapping.

## Understanding NUTS Levels in Turkey

Turkey's administrative geography follows the NUTS (Nomenclature of Territorial
Units for Statistics) system:

| Level | Turkish Term | English Translation | Count | Example |
|-------|--------------|---------------------|------:|---------|
| **NUTS-2** | İstatistiki Bölge | Statistical Region | 26 | İstanbul, Ankara |
| **NUTS-3** | İl | Province | 81 | İzmir, Antalya |
| **LAU-1** | İlçe | District | 973 | Kadıköy, Çankaya |
| **Level 9** | Yerleşim Noktası | Settlement Point | 1,003 | City centers |

Levels 2, 3, and 4 return `MULTIPOLYGON` geometries. Level 9 returns `POINT`
geometry.

## Setup

```{r setup}
library(tuikr)
library(dplyr)
library(ggplot2)
library(sf)
library(stringr)
```

## Discovering Available Variables

Start by exploring the available geographic indicators:

```{r variables}
variables <- geo_data()

dplyr::glimpse(variables)

variables |>
  dplyr::filter(stringr::str_detect(var_name, "Okul|Eğitim"))
```

`geo_data()` without parameters returns metadata. The `var_levels` column is a
list-column, where each element is an integer vector of supported geographic
levels.

## Downloading Geographic Data

### Basic Download

To download values for a specific variable:

```{r download}
cinema_data <- geo_data(
  variable_level = 3,
  variable_no = "SNM-GK160951-O33303",
  variable_source = "medas",
  variable_period = "yillik",
  variable_recnum = 5
)

head(cinema_data)
```

### Finding Variable Parameters

Extract the required parameters from the metadata table:

```{r params}
var_info <- variables |>
  dplyr::filter(var_num == "SNM-GK160951-O33303")

var_info |>
  dplyr::select(var_num, var_source, var_period, var_recordnum, var_levels)

var_info$var_levels[[1]]
```

`var_levels` is a list-column. `var_info$var_levels[[1]]` returns the integer
vector of supported levels for that variable.

### Level Availability

Not every indicator is published at every geographic level. An unavailable
request errors:

```{r level_error, error = TRUE, eval = identical(Sys.getenv("IN_PKGDOWN"), "true")}
geo_data(
  variable_level = 4,
  variable_no = "TFE-GK105747-O23001",
  variable_source = "medas",
  variable_period = "yillik",
  variable_recnum = 5
)
```

Check `var_levels` before you download. A small helper makes that explicit:

```{r helper}
download_if_available <- function(variable_no,
                                  variable_level,
                                  variable_source,
                                  variable_period,
                                  variable_recnum) {
  variable_metadata <- geo_data()

  available_levels <- variable_metadata |>
    dplyr::filter(var_num == variable_no) |>
    dplyr::pull(var_levels) |>
    purrr::pluck(1)

  if (variable_level %in% available_levels) {
    return(
      geo_data(
        variable_no = variable_no,
        variable_level = variable_level,
        variable_source = variable_source,
        variable_period = variable_period,
        variable_recnum = variable_recnum
      )
    )
  }

  stop(
    sprintf(
      "Level %s is not available for %s. Available levels: %s",
      variable_level,
      variable_no,
      paste(available_levels, collapse = ", ")
    )
  )
}
```

## Downloading Map Boundaries

Download the matching geometry with `geo_map()`:

```{r boundaries}
nuts2 <- geo_map(2)
nuts3 <- geo_map(3)
lau1 <- geo_map(4)
settlements <- geo_map(9)
```

`geo_map()` returns `sf` objects in WGS 84 (EPSG:4326). For levels 2, 3, and
4, the `code` column is ready for `dplyr::left_join()` with data returned by
`geo_data()`.

## Creating Choropleth Maps

### NUTS-2 Example: Laying Hens

```{r nuts2_map}
laying_hens <- geo_data(
  variable_level = 2,
  variable_no = "HYV-GK1696800-O32507",
  variable_source = "medas",
  variable_period = "yillik",
  variable_recnum = 20
) |>
  dplyr::filter(date %in% c("2015", "2023")) |>
  dplyr::mutate(
    yumurta_tavugu_sayisi_adet = as.numeric(yumurta_tavugu_sayisi_adet),
    date = as.integer(date)
  )

geo_map(2) |>
  dplyr::left_join(laying_hens, by = "code") |>
  ggplot2::ggplot() +
  ggplot2::geom_sf(ggplot2::aes(fill = yumurta_tavugu_sayisi_adet), color = "white") +
  ggplot2::coord_sf(datum = NA) +
  ggplot2::facet_wrap(~date, ncol = 2) +
  ggplot2::scale_fill_viridis_c(option = "plasma") +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "bottom",
    legend.key.width = grid::unit(3, "cm")
  ) +
  ggplot2::labs(
    title = "Laying Hens (2015 vs 2023)",
    fill = "Count",
    x = "Longitude (degrees)",
    y = "Latitude (degrees)",
    caption = "Source: TUIK"
  )
```

### NUTS-3 Example: Housing Sales

```{r nuts3_map}
housing_sales <- geo_data(
  variable_level = 3,
  variable_no = "INS-GK055-O006",
  variable_source = "ilGostergeleri",
  variable_period = "yillik",
  variable_recnum = 5
) |>
  dplyr::filter(date == "2025") |>
  dplyr::mutate(konut_satis_sayilari_toplam = as.numeric(konut_satis_sayilari_toplam))

geo_map(3) |>
  dplyr::left_join(housing_sales, by = "code") |>
  ggplot2::ggplot() +
  ggplot2::geom_sf(ggplot2::aes(fill = konut_satis_sayilari_toplam), color = "white", linewidth = 0.1) +
  ggplot2::coord_sf(datum = NA) +
  ggplot2::scale_fill_viridis_c(option = "viridis") +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "bottom",
    legend.key.width = grid::unit(3, "cm")
  ) +
  ggplot2::labs(
    title = "Housing Sales by Province (2025)",
    fill = "Sales",
    x = "Longitude (degrees)",
    y = "Latitude (degrees)",
    caption = "Source: TUIK"
  )
```

### LAU-1 Example: Illiterate Population

```{r lau1_map}
illiterate_population <- geo_data(
  variable_level = 4,
  variable_no = "ULE-GK160887-O29502",
  variable_source = "medas",
  variable_period = "yillik",
  variable_recnum = 5
) |>
  dplyr::filter(date == "2022") |>
  dplyr::mutate(okuma_yazma_bilmeyen_sayisi = as.numeric(okuma_yazma_bilmeyen_sayisi))

geo_map(4) |>
  dplyr::left_join(illiterate_population, by = "code") |>
  ggplot2::ggplot() +
  ggplot2::geom_sf(ggplot2::aes(fill = okuma_yazma_bilmeyen_sayisi), linewidth = 0.1) +
  ggplot2::coord_sf(datum = NA) +
  ggplot2::scale_fill_viridis_c(option = "inferno") +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "bottom",
    legend.key.width = grid::unit(3, "cm")
  ) +
  ggplot2::labs(
    title = "Illiterate Population by District (2022)",
    fill = "Count",
    x = "Longitude (degrees)",
    y = "Latitude (degrees)",
    caption = "Source: TUIK"
  )
```

### Regional Focus: Ankara Region

```{r regional}
district_population <- geo_data(
  variable_level = 4,
  variable_no = "ADNKS-GK137473-O29001",
  variable_source = "medas",
  variable_period = "yillik",
  variable_recnum = 5
) |>
  dplyr::filter(date == "2025") |>
  dplyr::mutate(toplam_nufus = as.numeric(toplam_nufus))

geo_map(4) |>
  dplyr::left_join(district_population, by = "code") |>
  dplyr::filter(bolgeKodu == "TR51") |>
  ggplot2::ggplot() +
  ggplot2::geom_sf(ggplot2::aes(fill = toplam_nufus), color = "white", linewidth = 0.1) +
  ggplot2::coord_sf(datum = NA) +
  ggplot2::scale_fill_viridis_c(option = "viridis") +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "bottom",
    legend.key.width = grid::unit(3, "cm")
  ) +
  ggplot2::labs(
    title = "Ankara Region: District Population (2025)",
    fill = "Population",
    x = "Longitude (degrees)",
    y = "Latitude (degrees)",
    caption = "Source: TUIK"
  )
```

## Advanced Cartography

`geogrid` and `cartogram` are listed in `Suggests`. Install them separately if
you want to run the cartogram examples:

```r
install.packages(c("geogrid", "cartogram"))
```

### Hexagonal Cartograms

```{r hex, message = FALSE}
library(geogrid)

turkey_hex_map <- sf::st_transform(geo_map(3), crs = 3395)

automobile_data <- geo_data(
  variable_level = 3,
  variable_no = "ULS-GK093-O009",
  variable_source = "ilGostergeleri",
  variable_period = "yillik",
  variable_recnum = 5
) |>
  dplyr::filter(date == "2024") |>
  dplyr::mutate(bin_kisi_basina_otomobil_sayisi = as.numeric(bin_kisi_basina_otomobil_sayisi))

hex_input <- dplyr::left_join(turkey_hex_map, automobile_data, by = "code")

hex_grid <- geogrid::calculate_grid(
  shape = hex_input,
  grid_type = "hexagonal",
  seed = 9
)

hex_map <- geogrid::assign_polygons(hex_input, hex_grid)

hex_map |>
  dplyr::mutate(
    name = stringr::str_replace(name, "AFYONKARAHİSAR", "AFYON"),
    name = stringr::str_replace(name, "KAHRAMANMARAŞ", "K.MARAŞ")
  ) |>
  ggplot2::ggplot() +
  ggplot2::geom_sf(ggplot2::aes(fill = bin_kisi_basina_otomobil_sayisi), linewidth = 0.1) +
  ggplot2::geom_sf_text(ggplot2::aes(label = name), color = "black", size = 2) +
  ggplot2::coord_sf(datum = NA) +
  ggplot2::scale_fill_viridis_c(option = "plasma") +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "bottom",
    legend.key.width = grid::unit(3, "cm")
  ) +
  ggplot2::labs(
    title = "Automobiles per 1,000 People by Province (2024)",
    fill = "Per 1,000",
    x = "Easting (m)",
    y = "Northing (m)",
    caption = "Source: TUIK"
  )
```

### Dorling Cartograms

```{r dorling, message = FALSE}
library(cartogram)

district_map <- sf::st_transform(geo_map(4), crs = 3395)

district_population <- geo_data(
  variable_level = 4,
  variable_no = "ADNKS-GK137473-O29001",
  variable_source = "medas",
  variable_period = "yillik",
  variable_recnum = 5
) |>
  dplyr::filter(date == "2024") |>
  dplyr::mutate(toplam_nufus = as.numeric(toplam_nufus))

district_illiteracy <- geo_data(
  variable_level = 4,
  variable_no = "ULE-GK160887-O29502",
  variable_source = "medas",
  variable_period = "yillik",
  variable_recnum = 5
) |>
  dplyr::filter(date == "2024") |>
  dplyr::mutate(okuma_yazma_bilmeyen_sayisi = as.numeric(okuma_yazma_bilmeyen_sayisi))

district_metrics <- dplyr::left_join(
  district_population,
  district_illiteracy,
  by = c("code", "date")
) |>
  dplyr::mutate(illiteracy_rate = 100 * (okuma_yazma_bilmeyen_sayisi / toplam_nufus))

district_metrics <- dplyr::left_join(district_map, district_metrics, by = "code")

dorling_map <- cartogram::cartogram_dorling(
  district_metrics,
  "okuma_yazma_bilmeyen_sayisi",
  0.4
)

dorling_map |>
  ggplot2::ggplot() +
  ggplot2::geom_sf(ggplot2::aes(fill = illiteracy_rate), linewidth = 0.1) +
  ggplot2::coord_sf(datum = NA) +
  ggplot2::scale_fill_viridis_c(option = "inferno") +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    legend.position = "bottom",
    legend.key.width = grid::unit(3, "cm")
  ) +
  ggplot2::labs(
    title = "Illiteracy Rate by District (2024)",
    subtitle = "Circle size = total count",
    fill = "Rate (%)",
    x = "Easting (m)",
    y = "Northing (m)",
    caption = "Source: TUIK"
  )
```
