---
title: "Getting Started with tuikr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with tuikr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
build_examples <- identical(Sys.getenv("IN_PKGDOWN"), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = build_examples
)
```

## Introduction

`tuikr` provides access to Turkish Statistical Institute (TUIK) data from two
portals:

1. **Main portal** ([veriportali.tuik.gov.tr](https://veriportali.tuik.gov.tr/)):
   Statistical themes, tables, and databases through SDMX and portal resources
2. **Geographic portal** ([cip.tuik.gov.tr](https://cip.tuik.gov.tr/)):
   Spatial data and ready-to-map geometries

## Installation

Install the development version from GitHub:

```r
# install.packages("devtools")
devtools::install_github("emraher/tuikr")
```

Load the package:

```{r setup}
library(tuikr)
```

## Statistical Data Workflow

### Step 1: Discover Themes

TUIK organizes portal content into theme folders. Start by listing the themes:

```{r themes}
themes <- statistical_themes(lang = "en")
themes
```

Theme IDs are character strings. Pass `"11"` rather than `11` so your code
matches the values returned by `statistical_themes()`.

### Step 2: Search Tables

Theme `"11"` is Population and Demography. Use it to list SDMX-backed datasets
and downloadable files:

```{r tables}
population_tables <- statistical_tables("11", lang = "en")
head(population_tables)
```

The `node_type` column tells you how to use each row:

- `"dataflow"`: SDMX-backed. Use `statistical_data()` with the `dataflow_id`
  column.
- `"istab"`: Direct file download. Use the `table_url` column with
  `download.file()`.

### Step 3: Find Database URLs

Use the same theme ID to retrieve legacy interactive database links:

```{r databases}
population_databases <- statistical_databases("11", lang = "en")
head(population_databases)
```

These URLs point to the legacy Biruni interface. They are browser targets, not
downloadable files.

### Step 4: Download SDMX Data

Filter to SDMX rows, then request one dataflow:

```{r sdmx}
pop_flows <- dplyr::filter(population_tables, node_type == "dataflow")

pop_data <- statistical_data(
  dataflow_id = pop_flows$dataflow_id[1],
  start = "2020",
  end = "2023"
)
dplyr::glimpse(pop_data)
```

The output is always long format. When TUIK exposes code-list metadata, coded
dimensions appear beside adjacent `*_label` columns with human-readable labels.

### Step 5: Download File Tables

Use `istab` rows when you want the portal's downloadable spreadsheets:

```{r download}
population_file_tables <- dplyr::filter(population_tables, node_type == "istab")

xls_path <- tempfile(fileext = ".xls")
download.file(
  population_file_tables$table_url[1],
  destfile = xls_path,
  mode = "wb"
)

table_data <- readxl::read_xls(xls_path)
head(table_data)
```

TUIK Excel files often have multi-row headers, footer notes, and mixed
Turkish-English labels. See the Known Issues vignette for cleaning strategies.

### Step 6: Browse Other Portal Resources

`statistical_resources()` returns the broader portal catalog for a theme:

```{r resources}
# Returns all portal resources for the theme
pop_resources <- statistical_resources("11")
dplyr::count(pop_resources, resource_type)
```

Use this when you want more than tables and databases and need a single view of
the theme's portal resources.

## Validation

The theme helpers validate their inputs before making a request:

```{r validate_invalid_theme, error = TRUE, eval = identical(Sys.getenv("IN_PKGDOWN"), "true")}
statistical_tables("aaa", lang = "en")
```

```{r validate_multiple_themes, error = TRUE, eval = identical(Sys.getenv("IN_PKGDOWN"), "true")}
statistical_tables(c("11", "12"), lang = "en")
```

## Troubleshooting

### Network Issues

Check TUIK portal accessibility. See
[Issue #2](https://github.com/emraher/tuikr/issues/2) for known connection
problems.

### SDMX Key Errors

If `statistical_data()` errors with a message about unconstrained dimensions,
supply a narrower SDMX key with the `key` argument. The key uses dot notation,
for example `key = "A.B.ALL"`.

### Website Structure Changes

TUIK portal updates can break scraping or resource discovery. Check for package
updates and report issues on
[GitHub](https://github.com/emraher/tuikr/issues).
