---
title: "Getting Started with tuikr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with tuikr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
build_examples <- identical(Sys.getenv("IN_PKGDOWN"), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = build_examples
)
```

## Introduction

`tuikr` provides access to Turkish Statistical Institute (TUIK) data from two portals:

1. **Main portal** ([veriportali.tuik.gov.tr](https://veriportali.tuik.gov.tr/)): Statistical themes, tables, and databases via a JSON API and SDMX dataflows
2. **Geographic portal** ([cip.tuik.gov.tr](https://cip.tuik.gov.tr/)): Spatial data and maps

## Installation

Install the development version from GitHub:

```{r install, eval=FALSE}
# install.packages("devtools")
devtools::install_github("emraher/tuikr")
```

Load the package:

```{r setup}
library(tuikr)
library(tidyverse)
```

## Statistical Data Workflow

### Step 1: Discover Themes

TUIK organizes data into thematic categories. Start by listing all available themes:

```{r themes}
(themes <- statistical_themes())
```

Returns a tibble with theme names (Turkish) and IDs.

### Step 2: Find Tables

Retrieve data tables and SDMX dataflows for a theme:

```{r tables}
# Get tables for Justice and Elections (theme 1)
justice_tables <- statistical_tables("1")

# View the first few tables
head(justice_tables)
```

Returns a tibble with `node_type` column: `"dataflow"` for SDMX interactive datasets,
`"istab"` for direct file downloads. SDMX dataflows include a `dataflow_id`,
which is the canonical machine identifier used by `statistical_data()`.

### Step 3: Discover All Portal Resources

If you want more than tables, use the broader resource catalog:

```{r resources}
# Get every supported portal resource for the same theme
justice_resources <- statistical_resources("1")

# Keep only press releases and reports
justice_publications <- statistical_resources(
  "1",
  type = c("press", "report")
)
```

The function validates input:

```{r validate, error=TRUE, eval=FALSE}
# Invalid theme ID
statistical_tables("aaa")
#> Error: You should select a valid theme ID!
```

```{r validate2, error=TRUE, eval=FALSE}
# Multiple themes not allowed
statistical_tables(c("1", "2"))
#> Error: You can select only one theme!
```

### Step 4: Get Database URLs

Retrieve legacy interactive database URLs:

```{r databases}
justice_dbs <- statistical_databases("1")
```


## Accessing Data

### SDMX Dataflows

Access SDMX-backed datasets through the exported download helpers:

```{r sdmx}
# Filter to SDMX dataflows
dataflows <- dplyr::filter(justice_tables, node_type == "dataflow")

# Download observations
justice_data <- statistical_data(
  dataflow_id = dataflows$dataflow_id[1]
)

# Return the browser's default table layout instead of long observations
birth_table <- statistical_data(
  dataflow_id = "TR,DF_DOGUM_IL_YASA_OZEL_DOGHIZ,1.0",
  layout = "table",
  lang = "en"
)
```

`statistical_data()` uses the same `dataflow_id` and defaults to
`key = "ALL"`, but some datasets require a more specific SDMX key. Use
`layout = "table"` when you want the browser's default pivoted table rather
than the long SDMX observation format.

### File Downloads

Download Excel files directly (`istab` nodes):

```{r download}
# Filter to file downloads
file_tables <- dplyr::filter(justice_tables, node_type == "istab")

# Download a specific file
download.file(
  file_tables$table_url[1],
  destfile = "/tmp/tuik_data.xls",
  mode = "wb"
)

# Read into R
library(readxl)
table_data <- read_xls("/tmp/tuik_data.xls")
```

TUIK Excel files often need cleaning: multiple headers, footer notes, mixed languages. Use `skip`, `n_max`, and `janitor::clean_names()`.

## Troubleshooting

### Network Issues

Check TUIK portal accessibility. See [Issue #2](https://github.com/emraher/tuikr/issues/2) for known connection problems.

### Website Structure Changes

TUIK website updates can break scraping. Check for package updates and report issues on [GitHub](https://github.com/emraher/tuikr/issues).

### Excel File Problems

Try `readxl::read_excel()`, use `skip` and `n_max` parameters, or inspect files manually.
