---
title: "Downloading Statistical Data with tuikr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Downloading Statistical Data with tuikr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
build_examples <- identical(Sys.getenv("IN_PKGDOWN"), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = build_examples,
  message = FALSE,
  warning = FALSE
)
```

## Introduction

`tuikr` exposes two statistical download paths from the TUIK portal:

- `dataflow` rows point to SDMX datasets that work with `tuikr::statistical_data()`
- `istab` rows point to spreadsheet downloads served by the portal

This vignette walks through one theme, downloads one SDMX series, and then
downloads one spreadsheet file with the same browser-style session that the
portal expects.

## Discover a Theme

Start with the theme catalog and locate the Population and Demography folder:

```{r themes}
theme_catalog <- tuikr::statistical_themes()

theme_catalog |>
  dplyr::filter(theme_id == "11")
```

## Inspect Table Nodes

`tuikr::statistical_tables()` returns both SDMX dataflows and spreadsheet
downloads for the same theme. The `node_type` column tells you which path you
are looking at.

```{r tables}
population_tables <- tuikr::statistical_tables("11")

population_tables |>
  dplyr::count(node_type)
```

```{r table-preview}
population_tables |>
  dplyr::select(table_name, node_type, dataflow_id, table_url) |>
  dplyr::slice_head(n = 6)
```

## Check Legacy Database Links

The legacy database URLs are still useful for manual browsing, even though they
are not the main programmatic download path.

```{r databases}
population_databases <- tuikr::statistical_databases("11")

population_databases |>
  dplyr::select(db_name, db_url) |>
  dplyr::slice_head(n = 3)
```

## Download an SDMX Dataset

Use the `dataflow_id` to request observations directly from the SDMX endpoint.
This chunk discovers the current flow ID for the "Age dependency ratio" table
before it calls `tuikr::statistical_data()`.

```{r sdmx-download}
age_dependency_flow <- population_tables |>
  dplyr::filter(
    node_type == "dataflow",
    table_name == "Age dependency ratio"
  ) |>
  dplyr::slice_head(n = 1)

population_data <- tuikr::statistical_data(
  dataflow_id = age_dependency_flow$dataflow_id[[1]],
  start = "2020",
  end = "2023"
)

population_data
```

`tuikr::statistical_data()` adds adjacent `*_label` columns when TUIK publishes
code-list labels, so you can work with readable values without losing the
original codes.

## Download a Spreadsheet Table

`istab` rows expose the spreadsheet download link in `table_url`.

```{r spreadsheet-download}
population_file_url <- population_tables |>
  dplyr::filter(node_type == "istab") |>
  dplyr::slice_head(n = 1) |>
  dplyr::pull(table_url)

population_file_url
```

Open that URL in a browser, or pass it to your own download workflow. `istab`
files often arrive with stacked headers, mixed labels, and source notes inside
the worksheet.

## Inspect the Full Resource Catalog

Use `tuikr::statistical_resources()` when you want one table of all supported
resource types for a theme.

```{r resources}
population_resources <- tuikr::statistical_resources("11")

population_resources |>
  dplyr::count(resource_type)
```
