---
title: "Getting Started with tuikr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with tuikr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
build_examples <- identical(Sys.getenv("IN_PKGDOWN"), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = build_examples
)
```

## Introduction

`tuikr` provides access to Turkish Statistical Institute (TUIK) data from two portals:

1. **Main portal** ([veriportali.tuik.gov.tr](https://veriportali.tuik.gov.tr/)): Statistical themes, tables, and databases via a JSON API and SDMX dataflows
2. **Geographic portal** ([cip.tuik.gov.tr](https://cip.tuik.gov.tr/)): Spatial data and maps

## Installation

Install the development version from GitHub:

```{r install, eval=FALSE}
# install.packages("devtools")
devtools::install_github("emraher/tuikr")
```

Load the package:

```{r setup}
library(tuikr)
```

## Statistical Data Workflow

### Step 1: Discover Themes

TUIK organizes data into thematic categories. Start by listing all available themes:

```{r themes}
(themes <- statistical_themes(lang = "en"))
```

Returns a tibble with theme names (English) and IDs.

### Step 2: Search Tables

For example, theme `"11"` is Population and Demography. Search that theme for
tables, SDMX dataflows, and direct file downloads:

```{r tables}
population_tables <- statistical_tables("11", lang = "en")
head(population_tables)
```

Returns a tibble with `node_type` column: `"dataflow"` for SDMX interactive datasets,
`"istab"` for direct file downloads. SDMX dataflows include a `dataflow_id`,
which is the canonical machine identifier used by `statistical_data()`.

### Step 3: Find Database URLs

Use the same theme ID to retrieve legacy interactive database links:

```{r databases}
population_databases <- statistical_databases("11", lang = "en")
head(population_databases)
```

These URLs point to the older Biruni query interface rather than downloadable
files.

### Step 4: Download SDMX Data

Use the `dataflow_id` values from `statistical_tables()` when you want machine-readable data:

```{r sdmx}
population_dataflows <- dplyr::filter(population_tables, node_type == "dataflow")


population_dataflows |> 
  dplyr::select(dataflow_id, table_name) |> 
  dplyr::slice(2)

population_data <- statistical_data(
  dataflow_id = population_dataflows$dataflow_id[2],
  start = "2023",
  end = "2024",
  lang = "en"
)

head(population_data)
```

`statistical_data()` defaults to `key = "ALL"`. Some datasets need a more
specific SDMX key.

In long output, coded dimensions keep their original machine-readable values
and gain adjacent `*_label` columns when TUIK exposes human-readable metadata
for those codes.

### Step 5: Download File Tables

Download Excel files directly from `istab` rows:

```{r download}
population_file_tables <- dplyr::filter(population_tables, node_type == "istab")

xls_path <- tempfile(fileext = ".xls")
download.file(
  population_file_tables$table_url[1],
  destfile = xls_path,
  mode = "wb"
)
table_data <- readxl::read_xls(xls_path)
head(table_data)
```

TUIK Excel files often need cleaning because they may contain multiple header
rows, footer notes, and mixed languages.

### Step 6: Browse Other Portal Resources

Use `statistical_resources()` when you want more than tables and databases:

```{r resources}
population_resources <- statistical_resources("11", lang = "en")
head(population_resources)

population_publications <- statistical_resources(
  "11",
  type = c("press", "report"),
  lang = "en"
)
head(population_publications)
```

The theme functions validate input:

```{r validate, error=TRUE, eval=FALSE}
# Invalid theme ID
statistical_tables("aaa", lang = "en")
#> Error: You should select a valid theme ID!
```

```{r validate2, error=TRUE, eval=FALSE}
# Multiple themes not allowed
statistical_tables(c("11", "12"), lang = "en")
#> Error: You can select only one theme!
```

## Troubleshooting

### Network Issues

Check TUIK portal accessibility. See [Issue #2](https://github.com/emraher/tuikr/issues/2) for known connection problems.

### Website Structure Changes

TUIK website updates can break scraping. Check for package updates and report issues on [GitHub](https://github.com/emraher/tuikr/issues).
