---
title: "Downloading Statistical Data with tuikr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Downloading Statistical Data with tuikr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
build_examples <- identical(Sys.getenv("IN_PKGDOWN"), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = build_examples,
  message = FALSE,
  warning = FALSE
)
```

## Introduction

`tuikr` exposes two statistical download paths from the TUIK portal:

- `dataflow` rows point to SDMX datasets that work with `tuikr::statistical_data()`
- `istab` rows point to spreadsheet downloads served by the portal

This vignette walks through one theme, downloads one SDMX series, and then
downloads one spreadsheet file with the same browser-style session that the
portal expects.

## Discover a Theme

Start with the theme catalog and locate the Population and Demography folder:

```{r themes}
theme_catalog <- tuikr::statistical_themes(lang = "en")

theme_catalog |>
  dplyr::filter(theme_id == "11")
```

## Inspect Table Nodes

`tuikr::statistical_tables()` returns both SDMX dataflows and spreadsheet
downloads for the same theme. The `node_type` column tells you which path you
are looking at.

```{r tables}
population_tables <- tuikr::statistical_tables("11", lang = "en")

population_tables |>
  dplyr::count(node_type)
```

```{r table-preview}
population_tables |>
  dplyr::select(table_name, node_type, dataflow_id, table_url) |>
  dplyr::slice_head(n = 6)
```

## Check Legacy Database Links

The legacy database URLs are still useful for manual browsing, even though they
are not the main programmatic download path.

```{r databases}
population_databases <- tuikr::statistical_databases("11", lang = "en")

population_databases |>
  dplyr::select(db_name, db_url) |>
  dplyr::slice_head(n = 3)
```

## Download an SDMX Dataset

Use the `dataflow_id` to request observations directly from the SDMX endpoint.
This chunk discovers the current flow ID for the "Age dependency ratio" table
before it calls `tuikr::statistical_data()`.

```{r sdmx-download}
age_dependency_flow <- population_tables |>
  dplyr::filter(
    node_type == "dataflow",
    table_name == "Age dependency ratio"
  ) |>
  dplyr::slice_head(n = 1)

population_data <- tuikr::statistical_data(
  dataflow_id = age_dependency_flow$dataflow_id[[1]],
  start = "2020",
  end = "2023",
  lang = "en"
)

population_data
```

`tuikr::statistical_data()` adds adjacent `*_label` columns when TUIK publishes
code-list labels, so you can work with readable values without losing the
original codes.

## Download a Spreadsheet Table

The spreadsheet endpoint rejects a cold scripted request. Reuse the portal
session with `crul::HttpClient` so the download works during the pkgdown build.

```{r spreadsheet-download}
browser_headers <- list(
  Accept = "application/json, text/plain, */*",
  `Accept-Language` = "en-US,en;q=0.9,tr-TR;q=0.8,tr;q=0.7",
  `User-Agent` = paste(
    "Mozilla/5.0 (Windows NT 10.0; Win64; x64)",
    "AppleWebKit/537.36 (KHTML, like Gecko)",
    "Chrome/122.0.0.0 Safari/537.36"
  )
)

portal_url <- "https://veriportali.tuik.gov.tr/en/statistical-themes"
cookie_store <- tempfile(fileext = ".txt")
request_opts <- list(cookiefile = cookie_store, cookiejar = cookie_store)

landing_client <- crul::HttpClient$new(
  url = portal_url,
  headers = browser_headers,
  opts = request_opts
)

landing_client$get()$raise_for_status()

population_file_tables <- population_tables |>
  dplyr::filter(node_type == "istab")

spreadsheet_client <- crul::HttpClient$new(
  url = population_file_tables$table_url[[1]],
  headers = c(
    browser_headers,
    list(
      Referer = portal_url,
      Origin = "https://veriportali.tuik.gov.tr"
    )
  ),
  opts = request_opts
)

spreadsheet_response <- spreadsheet_client$get()
spreadsheet_response$raise_for_status()

spreadsheet_path <- file.path(tempdir(), "tuikr-population-download")
writeBin(spreadsheet_response$content, spreadsheet_path)

spreadsheet_format <- readxl::excel_format(spreadsheet_path)
spreadsheet_format

spreadsheet_preview <- readxl::read_excel(
  spreadsheet_path,
  n_max = 10,
  col_names = FALSE
)

spreadsheet_preview
```

The preview is intentionally raw. `istab` files often arrive with stacked
headers, mixed labels, and source notes inside the worksheet.

## Inspect the Full Resource Catalog

Use `tuikr::statistical_resources()` when you want one table of all supported
resource types for a theme.

```{r resources}
population_resources <- tuikr::statistical_resources("11", lang = "en")

population_resources |>
  dplyr::count(resource_type)
```

## Troubleshooting

If an SDMX request complains about unconstrained dimensions, supply a narrower
`key` such as `key = "A.B.ALL"` to pin down the remaining dimensions.

If a spreadsheet download fails, keep the `table_url` but recreate the portal
session first. The file endpoint expects the same cookie-backed browser session
that the statistical portal uses for navigation.

If a workbook looks messy in `readxl`, treat that as upstream structure rather
than a parsing bug. `col_names = FALSE`, `skip`, and `janitor::clean_names()`
are the usual cleanup tools once you decide which rows hold the real table.
