---
title: "Known Issues and Limitations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Known Issues and Limitations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
build_examples <- identical(Sys.getenv("IN_PKGDOWN"), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = build_examples
)
```

## Introduction

`tuikr` depends on public TUIK portals. That makes the package useful, but it
also means a few limitations come from the upstream services rather than the R
package itself.

## TUIK Portal Changes

### The Challenge

TUIK migrated from `data.tuik.gov.tr` to `veriportali.tuik.gov.tr`. The current
package uses the newer portal, but the response structure can still change
without notice.

```{r website_change, error = TRUE}
statistical_databases("1")
```

### Workarounds

Check for package updates, report issues on
[GitHub](https://github.com/emraher/tuikr/issues), and keep in mind that one
resource category may break while others remain available.

## Messy Excel Files

### The Problem

TUIK Excel files often contain multi-row headers, merged cells, bilingual
labels, and source notes mixed into the data area.

### Example of a Messy File

```{r messy_example}
tables <- statistical_tables("1")
file_tables <- dplyr::filter(tables, node_type == "istab")

xls_path <- tempfile(fileext = ".xls")
download.file(file_tables$table_url[1], destfile = xls_path, mode = "wb")

raw_data <- readxl::read_xls(xls_path)
dplyr::glimpse(raw_data)
```

### Cleaning Strategies

```{r skip_rows}
clean_data <- readxl::read_xls(xls_path, skip = 3)
head(clean_data)
```

```{r limit_rows}
clean_data <- readxl::read_xls(xls_path, skip = 3, n_max = 10)
clean_data
```

```{r remove_na}
clean_data <- readxl::read_xls(xls_path, skip = 2) |>
  dplyr::filter(!is.na(...1)) |>
  dplyr::filter(!stringr::str_detect(...1, "Kaynak|Source"))

head(clean_data)
```

```{r clean_names}
clean_data <- readxl::read_xls(xls_path, skip = 3) |>
  janitor::clean_names() |>
  dplyr::rename(year = x1, total = x2)

head(clean_data)
```

Inspect a file before you automate the cleaning:

```{r manual_inspect}
dplyr::slice(raw_data, 1:5)
dplyr::slice(raw_data, (dplyr::n() - 4):dplyr::n())
```

## Network Dependencies

### Online-Only Functions

All exported functions call public TUIK services. Offline use is not supported.

### Handling Connection Failures

```{r connection_handling}
safely_get_themes <- function() {
  tryCatch(
    {
      return(statistical_themes())
    },
    error = function(error_condition) {
      message("Could not connect to the TUIK portal.")
      message("Error: ", error_condition$message)
      return(NULL)
    }
  )
}

themes <- safely_get_themes()

if (is.null(themes)) {
  stop("Cannot proceed without theme data.")
}
```

## API Limitations

### Rate Limiting

TUIK does not publish formal rate limits, but slower, sequential requests are
safer than tight loops.

```{r rate_limit}
theme_ids <- c("1", "2", "3")

all_tables <- purrr::map_dfr(theme_ids, function(theme_id) {
  Sys.sleep(1)
  statistical_tables(theme_id)
})
```

### SDMX Key Complexity

If `statistical_data()` returns an error like
`"Cannot construct a table layout because multiple unconstrained dimensions remain: obsTime"`,
the dataset needs a narrower SDMX key. The `key` argument uses dot notation,
where each dot separates a dimension. Use `"ALL"` for dimensions you do not
want to filter.

```r
# Default -- works for most datasets
statistical_data(dataflow_id = "DF_TUIK_001")

# Narrower key -- required for multi-dimensional datasets
statistical_data(dataflow_id = "DF_TUIK_001", key = "A.B.ALL")
```

Consult the TUIK SDMX browser at
[veriportali.tuik.gov.tr](https://veriportali.tuik.gov.tr/) to inspect the
dimension structure before you build a custom key.

### Data Availability by Level

Not all variables are available at all geographic levels. Check `var_levels`
before you download:

```{r check_levels}
download_if_available <- function(var_no,
                                  level,
                                  variable_source,
                                  variable_period,
                                  variable_recnum) {
  available_levels <- geo_data() |>
    dplyr::filter(var_num == var_no) |>
    dplyr::pull(var_levels) |>
    purrr::pluck(1)

  if (level %in% available_levels) {
    return(
      geo_data(
        variable_no = var_no,
        variable_level = level,
        variable_source = variable_source,
        variable_period = variable_period,
        variable_recnum = variable_recnum
      )
    )
  }

  message(sprintf("Level %d not available for %s", level, var_no))
  message(sprintf("Available levels: %s", paste(available_levels, collapse = ", ")))
  return(NULL)
}
```

### Wide Layout Removed

`statistical_data()` no longer accepts `layout = "wide"`. All output is long
format. If you have code that passed `layout = "wide"`, remove that argument.
The function now always returns long-format data with `*_label` columns
interleaved next to their coded counterparts.

## Locale and Date Parsing

### Geographic Data Dates

```{r parse_dates}
data <- geo_data(
  variable_no = "SNM-GK160951-O33303",
  variable_level = 3,
  variable_source = "medas",
  variable_period = "yillik",
  variable_recnum = 5
) |>
  dplyr::mutate(
    year = as.integer(substr(date, 1, 4)),
    month = dplyr::if_else(
      nchar(date) == 6,
      as.integer(substr(date, 5, 6)),
      NA_integer_
    )
  )
```

## GitHub Issue #2

[Issue #2](https://github.com/emraher/tuikr/issues/2) tracks connection
problems such as timeouts, SSL errors, and proxy failures.

## Character Encoding

```{r encoding}
Sys.getlocale("LC_CTYPE")
Sys.setlocale("LC_CTYPE", "en_US.UTF-8")
Sys.setlocale("LC_CTYPE", "Turkish_Turkey.utf8")
```

## Performance Considerations

### Large Downloads

Cache large downloads if you plan to reuse them:

```{r cache}
cache_file <- "data/district_population.rds"

if (file.exists(cache_file)) {
  district_data <- readRDS(cache_file)
} else {
  district_data <- geo_data(
    variable_level = 4,
    variable_no = "ADNKS-GK137473-O29001",
    variable_source = "medas",
    variable_period = "yillik",
    variable_recnum = 5
  )
  saveRDS(district_data, cache_file)
}
```

### Map Files

District geometries are comparatively large. Simplifying them can reduce memory
use when you only need small-scale maps:

```{r map_size}
lau1 <- geo_map(4)
format(object.size(lau1), units = "MB")

lau1_simple <- sf::st_simplify(lau1, dTolerance = 0.01)
format(object.size(lau1_simple), units = "MB")
```

## Reporting Issues

Report issues at
[github.com/emraher/tuikr/issues](https://github.com/emraher/tuikr/issues)
with the function call, error message, and `sessionInfo()`.

## Summary

| Issue | Severity | Workaround |
|-------|----------|------------|
| Website structure changes | High | Update package, report issues |
| Messy Excel files | Medium | Manual cleaning pipelines |
| Network dependency | Medium | Error handling, offline tests |
| SDMX key errors | Medium | Use `key` argument with dot-notation |
| Level availability | Low | Check metadata first |
| Rate limiting | Low | Add delays between requests |
| Turkish locales | Low | Set system locale |
